Cumulative link models for deep ordinal classification

### 2019-10-09 paper/elsarticle-template.tex
Paper
Correcciones paper

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

@@

\title{Deep ordinal classification based on cumulative link models}

@@

\begin{abstract}
	This paper proposes a deep convolutional neural network model for ordinal regression by considering a family of probabilistic ordinal link functions in the output layer. The link functions are those used for cumulative link models, which are traditional statistical linear models based on projecting each pattern into a 1-dimensional space. A set of ordered thresholds splits this space into the different classes of the problem. In our case, the projections are estimated by a non-linear deep neural network. To further improve the results, we combine these ordinal models with a loss function that takes into account the distance between the categories, based on the weighted Kappa index. Three different link functions are studied in the experimental study, and the results are contrasted with statistical analysis. The experiments run over two different ordinal classification problems, and the statistical tests confirm that these models improve the results of a nominal model and outperform other proposals considered in the literature.
\end{abstract}

@@

\section{Introduction}
\label{sect:introduction}
Deep learning, introduced by Lecun~et al. \cite{lecun2015deep}, combines multiple machine learning techniques and allows computational models that are composed of numerous processing layers to learn representations of data with various levels of abstraction. These methods have dramatically improved the state-of-the-art in many domains, such as image classification~\cite{yu2017convolutional,fang2019image,zhu2019image}, speech recognition~\cite{song2018music,liu2018speech}, control problems~\cite{mnih2015human}, object detection~\cite{sun2018face,chu2018deep,olmos2018automatic}, privacy and security protection~\cite{tan2018deep,yuan2019adversarial}, recovery of human pose~\cite{hong2015multimodal}, semantic segmentation~\cite{li2019deep} and image retrieval~\cite{tzelepi2018deep,qayyum2017medical,bai2018optimization}. Convolutional Neural Networks (CNN) are one of the types of deep networks that are designed to process data that come in the form of multiple arrays. CNNs are appropriate for images, video, speech and audio processing, and they have been used extensively in the last years for automatic classification tasks~\cite{dong2014learning,ronneberger2015u}. For images, each colour channel is represented by a 2D array, and convolutional layers extract the main features from the pixels, and, after that, a fully connected layer classify every sample based on its extracted features. At the output of the CNN, a softmax function provides the probabilities of the set of classes predefined in the model for classification tasks. However, the softmax may not be the best option depending on the problem considered.

@@

@@

%\textbf{(Métrica ordinal (MAE))}
ALALI et al. \cite{alali2018multi} proposed a complex CNN architecture for solving Twitter Sentiment Classification as an ordinal problem. They checked that using average pooling preserves significant features that provide more expressiveness to ordinal scale. They didn't propose any method to include the ordinal information into the classifier, but they tried to find the best CNN model architecture based on an ordinal metric.

@@

De la Torre et al.~\cite{de2018weighted} proposed the use of a continuous version of the quadratic weighted kappa (QWK) metric as loss function for the optimization algorithm. They compared this cost function against the traditional log-loss function using three different datasets, including the Diabetic Retinopathy database as the most complex one. They proved that their function could improve the results as it reduces overfitting and training time. Also, they checked the importance of hyper-parameter tuning. Later, in 2019, they published another work \cite{de2019deep} related to the Diabetic Retinopathy dataset where they combined the QWK loss function with the use of images of a higher resolution and new dataset partitions with many more samples on the training split. They achieved a proper classification score. Still, they tested the classifier against a tiny testing set compared to the original one from the Kaggle competition (around $12\%$ of the samples for testing against $60\%$ on the initial dataset splits). Further comparisons with this work are not possible as they didn't provide the dataset splits along with the paper.

Rios et al. \cite{rios2017ordinal} presented a CNN model designed to handle ordinal regression tasks on psychiatric notes. They combined an ordinal loss function, a CNN model and conventional feature engineering. Also, they applied a technique called Locally Interpretable Model-agnostic Explanation (LIME) to make the non-linear model more interpretable.

@@

\begin{figure}[!t]
	\centering
	\includegraphics[width=3in]{img/linkfunctions.pdf}
	\caption{Different link functions commonly used for CLMs.}
	\label{fig:linkfunctions}
\end{figure}

@@

After every ELU activation function of the convolutional layers, Batch Normalization~\cite{ioffe2015batch} is applied. This method reduces the internal covariate shift by normalizing layer outputs. It allows us to use higher learning rates and be less careful about weight initialization. It also eliminates the need for using regularization techniques like Dropout.

At the output of the network, the CLM is used (see Section \ref{sect:ordinalproblem}). Also, a learnable parameter $\tau$ has been used to rescale the projections used by the CLM to make it more stable and guarantee the convergence in most cases. The following expression describes the transformation applied to these projections:
\begin{equation}
\nonumber
f(\mathbf{x}) = \frac{l(\mathbf{x})}{\tau}.
\end{equation}
where $\tau$ is optimized as a free parameter.

@@

In order to ease reproducibility, the source code is available in a public repository\footnote{\url{https://github.com/ayrna/deep-ordinal-clm}}.
\subsection{Factors}
In our study, three different factors are considered:
\begin{itemize}
	\item \textit{Learning rate} (LR, $\eta$). LR is one of the most critical hyper-parameters to tune for training deep neural networks. Optimal learning rate can vary depending on the dataset and the CNN architecture. Previous works have presented some techniques that adjust this parameter in order to achieve better performance~\cite{smith2017cyclical,senior2013empirical}. In this work, we consider three different values for the initial value of this parameter: $10^{-2}$, $10^{-3}$ and $10^{-4}$.
	\item \textit{Batch size} (BS). Batch size is also an important parameter as it controls the number of weight updates that are made on every epoch. It can affect the training time and the model performance. In this paper, we try three different batch sizes for each dataset. For the DR dataset, we use $5$, $10$ and $15$, while, for Adience, $64$, $128$ and $256$ images are used. We took the batch sizes that were used in \cite{de2018weighted} and \cite{beckham2017unimodal} as a reference, and we expand the range on both sides.
	\item \textit{Link function} (LF). Different link functions are used for the CLM at the last layer output: \texttt{logit}, \texttt{probit} and complementary log-log (see Section \ref{sect:ordinalproblem}).
\end{itemize}

@@

\begin{table}[!t]
	\caption{Tukey's test results for both datasets.}
	\label{table:Tukey}
	\centering
	\footnotesize
	\begin{tabular}{cccccc}
		\hline\hline
		&                   & \multicolumn{2}{c}{\textbf{DR}} & \multicolumn{2}{c}{\textbf{Adience}} \\ \hline
		LF         &        LF         & Mean diff. &        Sig.        & Mean diff. &          Sig.           \\ \hline
		\texttt{logit}   &  \texttt{probit}  &  $0.046$   &      $0.000$       &  $-0.002$  &         $0.011$         \\
		& \texttt{clog-log} &  $0.084$   &      $0.000$       &  $0.012$   &         $0.000$         \\
		\texttt{probit}  &  \texttt{logit}   &  $-0.046$  &      $0.000$       &  $0.002$   &         $0.011$         \\
		& \texttt{clog-log} &  $0.038$   &      $0.000$       &  $0.014$   &         $0.248$         \\
		\texttt{clog-log} &  \texttt{logit}   &  $-0.084$  &      $0.000$       &  $-0.012$  &         $0.000$         \\
		&  \texttt{probit}  &  $-0.038$  &      $0.000$       &  $-0.014$  &         $0.248$         \\ \hline\hline
		LR         &        LR         & Mean diff. &        Sig.        & Mean diff. &          Sig.           \\ \hline
		$10^{-2}$     &     $10^{-3}$     &  $-0.046$  &      $0.000$       &  $-0.073$  &         $0.000$         \\
		&     $10^{-4}$     &  $0.148$   &      $0.000$       &  $-0.044$  &         $0.000$         \\
		$10^{-3}$     &     $10^{-2}$     &  $0.046$   &      $0.000$       &  $0.073$   &         $0.000$         \\
		&     $10^{-4}$     &  $0.194$   &      $0.000$       &  $0.029$   &         $0.023$         \\
		$10^{-4}$     &     $10^{-2}$     &  $-0.148$  &      $0.000$       &  $0.044$   &         $0.000$         \\
		&     $10^{-3}$     &  $-0.194$  &      $0.000$       &  $-0.029$  &         $0.023$         \\ \hline\hline
		BS         &        BS         & Mean diff. &        Sig.        & Mean diff. &          Sig.           \\ \hline
		64         &        128        &  $-0.041$  &      $0.026$       &     -      &            -            \\
		&        256        &  $-0.027$  &      $0.000$       &     -      &            -            \\
		128        &        64         &  $0.041$   &      $0.026$       &     -      &            -            \\
		&        256        &  $0.014$   &      $0.000$       &     -      &            -            \\
		256        &        64         &  $0.027$   &      $0.000$       &     -      &            -            \\
		&        128        &  $-0.014$  &      $0.000$       &     -      &            -\\\hline\hline
	\end{tabular}
\end{table}

@@

\section{Conclusions}
\label{sect:conclusions} 
This paper introduces a new deep ordinal network based on combining CLM models with a continuous QWK loss function. The proposed model is able to improve the performance of the deep network compared to the equivalent nominal version and other models proposed in previous works. Also, it is able to reduce the chance that the model gets stuck when training with some parameter configurations. We conclude that the optimal values for the different parameters considered are problem-dependant. The results highlight the importance of making an experimental design where all of these parameters are adjusted for each problem. In summary, the most significant contributions of the model proposal are the performance increase, the reduction of the number of parameters configurations that should be tried to find the best one and the prevention from over-fitting and stagnation.

@@

\section*{Acknowledgment}
This work has been partially subsidised by the TIN2017-85887-C2-1-P and TIN2017-90567-REDT projects of the Spanish Ministry of Economy and Competitiveness (MINECO), and FEDER funds of the European Union.

>>>>

%% `Elsevier LaTeX' style
%\bibliographystyle{elsarticle-num}
\bibliographystyle{IEEEtran}
%%%%%%%%%%%%%%%%%%%%%%%

@@

\title{Cumulative link models for deep ordinal classification}

@@

\begin{abstract}
	This paper proposes a deep convolutional neural network model for ordinal regression by considering a family of probabilistic ordinal link functions in the output layer. The link functions are those used for cumulative link models, which are traditional statistical linear models based on projecting each pattern into a 1-dimensional space. A set of ordered thresholds splits this space into the different classes of the problem. In our case, the projections are estimated by a non-linear deep neural network. To further improve the results, we combine these ordinal models with a loss function that takes into account the distance between the categories, based on the weighted Kappa index. Three different link functions are studied in the experimental study, and the results are contrasted with statistical analysis. The experiments run over two different ordinal classification problems, and the statistical tests confirm that these models improve the results of a nominal model and outperform other robust proposals considered in the literature.
\end{abstract}

@@

\section{Introduction}
\label{sect:introduction}
Deep learning, introduced by Lecun~et al. \cite{lecun2015deep}, combines multiple machine learning techniques and allows computational models that are composed of numerous processing layers to learn representations of data with various levels of abstraction. These methods have dramatically improved the state-of-the-art in many domains, such as image classification~\cite{yu2017convolutional,fang2019image,zhu2019image}, speech recognition~\cite{song2018music,liu2018speech}, control problems~\cite{mnih2015human}, object detection~\cite{sun2018face,chu2018deep,olmos2018automatic}, privacy and security protection~\cite{tan2018deep,yuan2019adversarial}, recovery of human pose~\cite{hong2015multimodal}, semantic segmentation~\cite{li2019deep} and image retrieval~\cite{tzelepi2018deep,qayyum2017medical,bai2018optimization}. Convolutional Neural Networks (CNN) are one of the types of deep networks that are designed to process data that come in the form of multiple arrays. CNNs are appropriate for images, video, speech and audio processing, and they have been used extensively in the last years for automatic classification tasks~\cite{dong2014learning,ronneberger2015u}. For images, each colour channel is represented by a 2D array, and convolutional layers extract the main features from the pixels, and, after that, a fully connected layer classify every sample based on its extracted features. At the output of the CNN, a softmax function provides the probabilities of the set of classes predefined in the model for classification tasks. However, the softmax may not be the best option depending on the classification problem considered.

@@

The highlights of this work are to introduce the CLM for CNNs combined with the QWK loss function to achieve high classification performance and better stability than previous works in terms of standard deviation and stagnation problems.

@@

%\textbf{(Métrica ordinal (MAE))}
Alali et al. \cite{alali2018multi} proposed a complex CNN architecture for solving Twitter Sentiment Classification as an ordinal problem. They checked that using average pooling preserves significant features that provide more expressiveness to ordinal scale. They didn't propose any method to include the ordinal information into the classifier, but they tried to find the best CNN model architecture based on an ordinal metric.

@@

De la Torre et al.~\cite{de2018weighted} proposed the use of a continuous version of the quadratic weighted kappa (QWK) metric as loss function for the optimization algorithm. They compared this cost function against the traditional log-loss function using three different datasets, including the Diabetic Retinopathy database as the most complex one. They proved that their function could improve the results as it reduces overfitting and training time. Also, they checked the importance of hyper-parameter tuning. Later, in 2019, another work related to the Diabetic Retinopathy dataset was published \cite{de2019deep} where they combined the QWK loss function with the use of images of a higher resolution and new dataset partitions with many more samples on the training split. They achieved a proper classification score but tested the classifier against a tiny testing set compared to the original one from the Kaggle competition (around $12\%$ of the samples for testing against $60\%$ on the initial dataset splits). Further comparisons with this work are not possible as they didn't provide the dataset splits along with the paper.

Rios et al. \cite{rios2017ordinal} presented a CNN model designed to handle ordinal regression tasks on psychiatric notes. They combined an ordinal loss function, a CNN model and conventional feature extraction. Also, the authors applied a technique called Locally Interpretable Model-agnostic Explanation (LIME) to make the non-linear model more interpretable.

@@

\begin{figure}[!t]
	\centering
	\includegraphics[width=\columnwidth]{img/linkfunctions.pdf}
	\caption{Different link functions commonly used for CLMs.}
	\label{fig:linkfunctions}
\end{figure}

@@

After every ELU activation function of the convolutional layers, Batch Normalization~\cite{ioffe2015batch} is applied. This method reduces the internal covariate shift by normalizing layer outputs. As the authors stated \cite{ioffe2015batch}, it allows us to use higher learning rates and be less careful about weight initialization. In some cases, it also eliminates the need for using regularization techniques like Dropout.

At the output of the network, the CLM is used (see Section \ref{sect:ordinalproblem}). Also, a learnable parameter $\tau$ rescales the projections used by the CLM to make it more stable and guarantee the convergence in most cases. The following expression describes the transformation applied to these projections:
\begin{equation}
\nonumber
f(\mathbf{x}) = \frac{l(\mathbf{x})}{\tau}.
\end{equation}
where $\tau$ is optimized as a free parameter along with the parameters of the model.

@@

In order to ease reproducibility, the source code is available in a public repository\footnote{\url{https://github.com/ayrna/deep-ordinal-clm}}.
\subsection{Factors}
In our study, three different factors are considered:
\begin{itemize}
	\item \textit{Learning rate} (LR, $\eta$). LR is one of the most critical hyper-parameters to tune for training deep neural networks. Optimal learning rate can vary depending on the dataset and the CNN architecture. Previous works have presented some techniques that adjust this parameter in order to achieve better performance~\cite{smith2017cyclical,senior2013empirical}. In this work, we consider three different values for the initial value of this parameter: $10^{-4}$, $10^{-3}$ and $10^{-2}$.
	\item \textit{Batch size} (BS). Batch size is also an important parameter as it controls the number of weight updates that are made on every epoch. It can affect the training time and the model performance. In this paper, we try three different batch sizes for each dataset. For the DR dataset, we use $5$, $10$ and $15$, while, for Adience, $64$, $128$ and $256$ images are used. We took the batch sizes that were used in \cite{de2018weighted} and \cite{beckham2017unimodal} as a reference, and we expand the range on both sides.
	\item \textit{Link function} (LF). Different link functions are used for the CLM at the last layer output: \texttt{logit}, \texttt{probit} and complementary log-log (see Section \ref{sect:ordinalproblem}).
\end{itemize}

@@

\begin{table}[!t]
	\caption{Tukey's test results for both datasets.}
	\label{table:Tukey}
	\centering
	\footnotesize
	\begin{tabular}{cccccc}
		\hline\hline
		&                   & \multicolumn{2}{c}{\textbf{DR}} & \multicolumn{2}{c}{\textbf{Adience}} \\ \hline
		LF         &        LF         & Mean diff. &        P-val        & Mean diff. &          P-val           \\ \hline
		\texttt{logit}   &  \texttt{probit}  &  $0.046$   &      $0.000$       &  $-0.002$  &         $0.011$         \\
		& \texttt{clog-log} &  $0.084$   &      $0.000$       &  $0.012$   &         $0.000$         \\
		\texttt{probit}  &  \texttt{logit}   &  $-0.046$  &      $0.000$       &  $0.002$   &         $0.011$         \\
		& \texttt{clog-log} &  $0.038$   &      $0.000$       &  $0.014$   &         $0.248$         \\
		\texttt{clog-log} &  \texttt{logit}   &  $-0.084$  &      $0.000$       &  $-0.012$  &         $0.000$         \\
		&  \texttt{probit}  &  $-0.038$  &      $0.000$       &  $-0.014$  &         $0.248$         \\ \hline\hline
		LR         &        LR         & Mean diff. &        P-val        & Mean diff. &          P-val           \\ \hline
		$10^{-2}$     &     $10^{-3}$     &  $-0.046$  &      $0.000$       &  $-0.073$  &         $0.000$         \\
		&     $10^{-4}$     &  $0.148$   &      $0.000$       &  $-0.044$  &         $0.000$         \\
		$10^{-3}$     &     $10^{-2}$     &  $0.046$   &      $0.000$       &  $0.073$   &         $0.000$         \\
		&     $10^{-4}$     &  $0.194$   &      $0.000$       &  $0.029$   &         $0.023$         \\
		$10^{-4}$     &     $10^{-2}$     &  $-0.148$  &      $0.000$       &  $0.044$   &         $0.000$         \\
		&     $10^{-3}$     &  $-0.194$  &      $0.000$       &  $-0.029$  &         $0.023$         \\ \hline\hline
		BS         &        BS         & Mean diff. &        P-val        & Mean diff. &          P-val           \\ \hline
		64         &        128        &  $-0.041$  &      $0.026$       &     -      &            -            \\
		&        256        &  $-0.027$  &      $0.000$       &     -      &            -            \\
		128        &        64         &  $0.041$   &      $0.026$       &     -      &            -            \\
		&        256        &  $0.014$   &      $0.000$       &     -      &            -            \\
		256        &        64         &  $0.027$   &      $0.000$       &     -      &            -            \\
		&        128        &  $-0.014$  &      $0.000$       &     -      &            -\\\hline\hline
	\end{tabular}
\end{table}

@@

\section{Conclusions}
\label{sect:conclusions} 
This paper introduces a new deep ordinal network based on combining CLM models with a continuous QWK loss function. The proposed model is able to improve the performance of the deep network compared to the equivalent nominal version and other models proposed in previous works. Also, it is able to reduce the chance that the model gets stuck when training with some parameter configurations. We conclude that the optimal values for the different parameters considered are problem-dependant. The results highlight the importance of making an experimental design where all of these parameters are adjusted for each problem. In summary, the most significant contributions of the model proposal are the performance increase, the reduction of the number of parameters configurations that should be tried to find the best one and the prevention from over-fitting and stagnation, which gives our method high stability compared to other alternatives.

@@

\section*{Acknowledgment}
This work has been partially subsidised by the TIN2017-85887-C2-1-P project of the Spanish Ministry of Economy and Competitiveness (MINECO), and FEDER funds of the European Union. Víctor Manuel Vargas's research has been subsidized by the FPU Predoctoral Program of the Spanish Ministry of Education, Culture and Sport (MECD), grant reference FPU18/00358.
